A seamless, client-centric programming model for type safe web applications

### 2014-05-02 paper.tex
Remove defensive Haste-justification.

Haste is by no means the only JavaScript-targeting compiler for a purely
functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
JavaScript; the Fay\ \cite{fay} language was designed from the ground up to
target the web space using a subset of Haskell; and there exist solutions for
compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to JavaScript as
well.

@@

While several of the aforementioned compilers fullfil these criteria, the
choice between them becomes almost arbitrary. We chose to base our solution on
Haste as we, by virtue of its authorship, have an intimate knowledge of its
internal workings, strengths and weaknesses. Without doubt, others may see many
reasons to make a different choice.

\subsection{Design goals}

Haste is first and foremost intended to be a practical compiler; the ability
to produce code which is lean and fast enough to be used in the real world is
critical. To this end, we make several design choices and tradeoffs, some of
which would be viewed as detrimental in a more purist context.

One such tradeoff is the decision to make use of the facilities provided by the
web browser - most prominently the JavaScript engine's native garbage
collector - as far as possible rather than implementing parts of the GHC
runtime in JavaScript. In the same vein, a conscious effort is made to generate
as ``regular-looking'' code as possible. This results in large gains in code
compactness and execution speed - in our experiences, compiler Haste programs
are often an order of magnitude smaller and several times faster than an
equivalent program compiled with some other Haskell to JavaScript compiler.

In some rare cases, however, this means having to sacrifice compatibility with
low-level GHC features. Most notably, Haste does not support weak references
or pre-emptive multitasking, as these would require modifications to the
runtime system and generated code which confound the structure of the
resulting JavaScript and have a detrimental impact on code size and execution
speed.

@@

As previously mentioned, Haste offloads much of the heavy lifting of
compilation - parsing, type checking, intermediate code generation and many
optimizations - onto GHC, and takes over code generation after the STG
generation step, at the very end of the compilation process. STG\ \cite{stg} is
the last intermediate representation used by GHC before the final code
generation takes place and has several benefits for use as Haste's source
language:

@@

\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Should this model prove
insufficiently expressive, extending our existing model with an
\lstinline!onClient! counterpart to \lstinline!onServer! is a simple task.

@@

As this modification only affects the program's entry, point which is already
given ``special'' treatment in Haskell compilers, and may even be implemented
using the plugin mechanism of GHC, requiring no modification of the compiler
itself, we believe this to be a practical, non-intrusive way to remove this
limitation, and are planning on releasing an implementation which includes
this limitation.

>>>>

Haste is by no means the only JavaScript-targeting compiler for a purely
functional language. In particular, the GHC-based GHCJS\ \cite{ghcjs} and
UHC\ \cite{uhc} compilers are both capable of compiling standard Haskell into
JavaScript; the Fay\ \cite{fay} language was designed from the ground up to
target the web space using a subset of Haskell; and there exist solutions for
compiling Erlang\ \cite{jserlang} and Clean\ \cite{jsclean} to JavaScript as
well. While the aforementioned compilers are the ones most interesting for
purely functional programming, there exist a wealth of other
JavaScript-targeting compilers, for virtually any language.

@@

As several of the aforementioned compilers fullfil these criteria, the
choice between them becomes almost arbitrary. Indeed, as Haste.App is compiler
agnostic, this decision boils down to one's personal preference. We chose to
base our solution on Haste as we, by virtue of its authorship, have an intimate
knowledge of its internal workings, strengths and weaknesses.
Without doubt, others may see many reasons to make a different choice.

@@

Haste offloads much of the heavy lifting of compilation - parsing,
type checking, intermediate code generation and many optimizations - onto GHC,
and takes over code generation after the STG generation step, at the very end
of the compilation process. STG\ \cite{stg} is the last intermediate
representation used by GHC before the final code generation takes place and
has several benefits for use as Haste's source language:

@@

\paragraph{Client-centricity} Unlike Conductance and Duetto, our approach
takes a firm stand, regarding the client as the driver in the client-server
relationship with the server taking on the role of a passive computational
or storage resource. The server may thus not call back into the client at
arbitrary points but is instead limited to returning answers to client side
queries. This is clearly less flexible than the back-and-forth model of Sunroof
and Duetto or the shared variables of Conductance. However, we believe that
this restriction makes program flow easier to follow and comprehend. Like the
immutability of Haskell, this model gives programmers a not-so-subtle hint
as to how they may want to structure their programs. Extending our existing
model with an \lstinline!onClient! counterpart to \lstinline!onServer! would be
a simple task, but we are not quite convinced that there is value in doing so.

@@

As this modification only affects the program's entry, point which is already
given ``special'' treatment in Haskell compilers, and may even be implemented
using the plugin mechanism of GHC, requiring no modification of the compiler
itself, we believe this to be a practical, non-intrusive way to remove this
limitation.
